Variational Monte Carlo on bosonic systems

### 2018-01-23 project_1/article/article.tex
Fix notation and correct the gradient calculation

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        V_{\text{ext}}(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    trap. The two-body Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one body operator, $h$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}\nabla_i^2
        + V_{\text{ext}}(\vf{r}_i),
    \end{align}
    (we assume equal mass)
    and the two-body interaction operator, $w$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
    \end{align}
    where $a$ is the hard sphere of the particle. The trial wavefunction we will
    be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \bra{\vf{r}}\ket{\Psi_T}
        = \prod_i^N g(\alpha, \beta, \vf{r}_i)
        \prod_{i < j}^N f(a, \vf{r}_i, \vf{r}_j),
    \end{align}
    where $\alpha$ and $\beta$ are variational parameters and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single-particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and the correlation wavefunction, $f$, is
    \begin{align}
        f(a, \vf{r}_i, \vf{r}_j)
        &=
        \begin{cases}
            0 & |\vf{r}_i - \vf{r}_j| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_i - \vf{r}_j|}
            \Bigr) & |\vf{r}_i - \vf{r}_j| > a.
        \end{cases}
    \end{align}

@@

            We start by finding an analytical expression for $a = 0$ and $\beta
            = 1$, i.e., a non-interaction harmonic oscillator given by the
            wavefunction

@@

@@

                = \sum_i^N \Bigl(

@@

                \Bigr).

@@

            operator (applied to a function) becomes

@@

                    x & \text{one dimension}, \\

@@

>>>>

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        V_{\text{ext}}(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    trap. The two-body Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one body operator, $h$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}\nabla_i^2
        + V_{\text{ext}}(\vf{r}_i),
    \end{align}
    (we assume equal mass)
    and the two-body interaction operator, $w$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
    \end{align}
    where $a$ is the hard sphere of the particle. The trial wavefunction we will
    be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \bra{\vf{r}}\ket{\Psi_T}
        = \bra{\vf{r}}\ket{\Phi_T}
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_i^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
    \end{align}
    where $\alpha$ and $\beta$ are variational parameters and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\ket{\Phi_T}$ the \textit{Slater permanent} consisting of the $N$ first
    single particle wavefunctions, and the correlation wavefunction, $f$, given
    by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
    \end{align}

@@

            We start by finding an analytical expression for the local energy
            with a system where we set $a = 0$ and $\beta = 1$, i.e., a system
            of non-interacting harmonic oscillators.

@@

                = \bra{\vf{r}}\ket{\Phi_T}

@@

                = \sum_i^N \biggl(

@@

                \biggr).

@@

            operator (applied to a scalar function) becomes

@@

                    x^2 & \text{one dimension}, \\

@@

        \subsubsection{The full system}
            Moving to the full system allowing $\beta$ to vary and setting $a
            \neq 0$ we write the trial wavefunction as
            \begin{align}
                \bra{\vf{r}}\ket{\Psi_T}
                &=
                \bra{\vf{r}}\ket{\Phi_T}
                \exp\Biggl(
                    \sum_{j < k}^N u(r_{jk})
                \Biggr) \\
                &=
                \Biggl(
                    \prod_i^N \phi(\vf{r}_i)
                \Biggr)
                \exp\Biggl(
                    \sum_{j < k}^N u(r_{jk})
                \Biggr),
            \end{align}
            where $r_{jk} = |\vf{r}_j - \vf{r}_k|$ and
            \begin{align}
                u(r_{jk}) = \ln\bigl[f(a, \vf{r}_j, \vf{r}_k)\bigr],
            \end{align}
            we wish to compute the local energy. The gradient is thus given by
            \begin{widetext}
                \begin{align}
                    \bra{\vf{r}}\nabla_k\ket{\Psi_T}
                    &= \nabla_k\brac{
                        \bra{\vf{r}}\ket{\Phi_T}
                        \exp\Biggl(
                            \sum_{j < l}^N u(r_{jl})
                        \Biggr)
                    } \\
                    &=
                    \Biggl\{
                        \nabla_k
                        \prod_i^N\phi(\vf{r}_i)
                    \Biggr\}
                    \exp\Biggl(
                        \sum_{j < l}^N u(r_{jl})
                    \Biggr)
                    +
                    \bra{\vf{r}}\ket{\Phi_T}
                    \nabla_k
                    \exp\Biggl(
                        \sum_{j < l}^N u(r_{jl})
                    \Biggr) \\
                    &=
                    \Biggl\{
                        \nabla_k\phi(\vf{r}_k)
                        \prod_{i \neq k}^N \phi(\vf{r}_i)
                    \Biggr\}
                    \exp\Biggl(
                        \sum_{j < l}^N u(r_{jl})
                    \Biggr)
                    +
                    \bra{\vf{r}}\ket{\Psi_T}
                    \nabla_k
                    \Biggl(
                        \sum_{m < n}^N u(r_{mn})
                    \Biggr).
                \end{align}
            The last term is extracted from the chain rule and can be evaluated
            as
            \begin{align}
                \nabla_k
                    \sum_{m < n}^N u(r_{mn})
                &=
                \sum_{m = 1}^{k - 1}
                \nabla_k u(r_{mk})
                +
                \sum_{n = k + 1}^N
                \nabla_k u(r_{kn}) \\
                &=
                \sum_{m \neq k}^N\nabla_k u(r_{km}),
            \end{align}
            as $r_{ij} = r_{ji}$.  We are thus left with
            \begin{align}
                \bra{\vf{r}}\nabla_k\ket{\Psi_T}
                &=
                \Biggl\{
                    \nabla_k\phi(\vf{r}_k)
                    \prod_{i \neq k}^N \phi(\vf{r}_i)
                \Biggr\}
                \exp\Biggl(
                    \sum_{j < k}^N u(r_{jk})
                \Biggr)
                +
                \bra{\vf{r}}\ket{\Psi_T}
                \sum_{m \neq k}^N\nabla_k u(r_{km}).
            \end{align}
            \end{widetext}
